```{r eval=FALSE, echo=FALSE, message=FALSE}
library(knitr)
library(markdown)
library(Cairo)

knit("tutorial.rmd","tutorial.md")
markdownToHTML("tutorial.md","tutorial.html")
```

```{r echo=FALSE, message=FALSE}
opts_chunk$set(warning=FALSE, fig.width=6, fig.height=6, dev="CairoPNG", stop=TRUE)
```

## Running an epigenome-wide association study (EWAS) in `ewaff`

### Preparing R

We first load the `ewaff` library.
```{r}
library(ewaff)
```

We indicate how many processors are available for performing analyses.
If this is not set explicitly, then the default is to use only
a single processor.
```{r}
options(mc.cores=4)
```

### Load a dataset

We will in fact generate a random dataset for this demonstration.
```{r}
set.seed(20171031)
n <- 500 ## n samples
s <- 100  ## s features/cpg sites

## variable of interest and covariates
data <- data.frame(variable=c(rep("A",n/2), rep("B",n/2)),      ## variable of interest (two groups)
                   continuous=rnorm(n),                         ## continuous covariate
                   categorical=factor(sample(0:3,n,replace=T)), ## categorical covariate
                   batch=factor(sample(0:2, size=n, replace=T)))        ## 3 batches

## correlation of each cpg site with the variable of interest
r <- runif(s, min=-1, max=1)

## methylation matrix randomly generated
## with batch effects and associations with the variable
## of interest (rows=cpg sites, cols=samples)
methylation <- t(sapply(r, function(r) {
    v <- as.numeric(data$variable)
    b <- data$batch
    ## batch effect
    b <- rnorm(unique(b), mean=0, sd=0.2)[b]
    ## noise
    e <- rnorm(length(v), mean=0, sd=sqrt(1-r^2))
    ## mean
    m <- runif(1, min=0.2, max=0.8)
    ## signal with mean=0, sd=1
    y <- (r*scale(v) + b + e)
    ## signal with mean=m, sd such that signal is 0..1
    f <- runif(1, min=0, max=min(m,1-m)/max(abs(y)))
    y <- y*f + m
    ## return result
    y
}))
rownames(methylation) <- paste("s", 1:nrow(methylation), sep="")
colnames(methylation) <- paste("p", 1:ncol(methylation), sep="")

## specify cpg site locations
manifest <- data.frame(chr=c(rep(1,s/2), rep(2,s/2)), pos=sample(1:(150*s), s))
manifest <- manifest[order(manifest$chr, manifest$pos),]

## generate 10 likely outliers
outliers <- cbind(sample(1:nrow(methylation), 10, replace=T),
                  sample(1:ncol(methylation), 10, replace=T))
methylation[outliers] <- ifelse(rowMeans(methylation)[outliers[,1]] > 0.5, 0, 1)
```

### Handling outliers

We use the 'iqr' method for handling outliers.5B
The 'iqr' method sets methylation levels that are outside
3*IQR of a CpG site to NA.
```{r}
methylation <- ewaff.handle.outliers(methylation, method="iqr")[[1]]
```

We can check that most of the outliers added to the dataset in the simulation
were set to NA.
```{r}
## outliers added
outliers
## outliers identified
which(is.na(methylation), arr.ind=T)
```
Some 'outliers' were missed because they are not actually outliers.

### Running the EWAS

Now we run the EWAS.  Notice that in this example
we don't explicitly include `batch` as a covariate,
although that is possible.
Here we let surrogate variable analysis (SVA)
generate batch covariates.
```{r}
sites.ret <- ewaff.sites(methylation ~ variable + continuous + categorical,
                         variable.of.interest="variable",
                         methylation=methylation,
                         data=data,
                         generate.confounders="sva",
                         random.subset=1,
                         n.confounders=1,
                         method="glm")
```

We show the top 10 associations.
```{r}
top.idx <- order(sites.ret$table$p.value)[1:10]
sites.ret$table[top.idx,]
```

Just for interest, we see if SVA detected batch.
```{r}
fit <- lm(sites.ret$design[,"sv1"] ~ data$batch)
coef(summary(fit))
```
It does seem like it did.

### Generating an EWAS report

```{r}
sum.ret <- ewaff.summary(sites.ret, manifest$chr, manifest$pos, methylation,
                         selected.cpg.sites="s58")

ewaff.report(sum.ret, output.file="output/report.html",
             author="Dom Rand",
             study="Associations in my kind of data")
```

### Other kinds of EWAS

#### Methylation is not the outcome
Methylation does not have to be the outcome variable.
In fact, any valid GLM model is possible.
In the following example, we make our binary variable the outcome.
```{r}
log.ret <- ewaff.sites(variable ~ methylation + continuous + categorical,
                       variable.of.interest="variable",
                       methylation=methylation,
                       data=data,
                       family="binomial",
                       generate.confounders="sva",
                       random.subset=1,
                       n.confounders=1,
                       method="glm")                     
```

Associations are identical to
those identified when methylation was the outcome.
```{r}
table(sites.ret$table$p.adjust < 0.05, log.ret$table$p.adjust < 0.05)
```


#### Variable of interest is complex

The variable of interest may be categorical with more than two categories.
```{r}
cats.ret <- ewaff.sites(methylation ~ categorical + variable + continuous,
                        variable.of.interest="categorical",
                        methylation=methylation,
                        data=data,
                        generate.confounders="sva",
                        random.subset=1,
                        n.confounders=1,
                        method="limma")
```

In this case, an f-statistic and p-value is calculated for the variable
along with statistics each binary 'dummy' variable.
```{r}
cats.ret$table[1:2,]
```

The variable of interest may actuually be multiple variables.
```{r}
vars.ret <- ewaff.sites(methylation ~ categorical + variable + continuous,
                        variable.of.interest=c("continuous","variable"),
                        methylation=methylation,
                        data=data,
                        generate.confounders="sva",
                        random.subset=1,
                        n.confounders=1,
                        method="limma")
```

Here again, an f-statistic is calcualted along with individual statistics
for each variable.
```{r}
vars.ret$table[1:5,]
```
